Enforced Invariant Confluence

### 2018-09-20 doc/paper/sections/abstract.tex
Integrating Joe's feedback.
Discussed limitations + shrunk paper.
From reviewer 11:

> Discussion of limitations and practical use can be made clearer.

From reviewer 12:

> While the ideas of the paper are very nice, the paper may more
> concretely describe how practical the algorithms are. For example, the
> first decision procedure (Algorithm 1) involves human to augment NR.
> In general, what are the tricky classes of transactions that may
> require lot of human interactions? Similar comments apply to segmented
> invariant confluence (Algorithm 5). What classes of
> transactions/applicatio

\begin{abstract}
Strongly consistent distributed systems are easy to reason about but face fundamental
limitations in availability and performance. Weakly consistent systems can be
implemented with very high performance but place a burden on the application
developer to reason about complex interleavings of execution.
\Invariantconfluence{} provides a formal framework for understanding when we
can get the best of both worlds. An \invariantconfluent{} object can be
efficiently replicated with no coordination needed to preserve its invariants.
However, actually determining whether or not an object is \invariantconfluent{}
is challenging: undecidable in general. In this paper, we establish conditions under
which a commonly used sufficient condition for \invariantconfluence{} is both
necessary and sufficient and use this condition to design (a) a general-purpose
interactive \invariantconfluence{} decision procedure and (b) a novel
sufficient condition that can be checked automatically. We then take a step
beyond \invariantconfluence{} and introduce a generalization of
\invariantconfluence{} called segmented \invariantconfluence{}, which allows us
to replicate non-\invariantconfluent{} objects with a small amount of
coordination.

>>>>

\begin{abstract}
Strongly consistent distributed systems are easy to reason about but face
fundamental limitations in availability and performance. Weakly consistent
systems can be implemented with very high performance but place a burden on the
application developer to reason about complex interleavings of execution.
\Invariantconfluence{} provides a formal framework for understanding when we
can get the best of both worlds. An \invariantconfluent{} object can be
efficiently replicated with no coordination needed to preserve its invariants.
However, actually determining whether or not an object is \invariantconfluent{}
is challenging: undecidable in general.

In this paper, we establish conditions under which a commonly used sufficient
condition for \invariantconfluence{} is both necessary and sufficient and use
this condition to design (a) a general-purpose interactive
\invariantconfluence{} decision procedure and (b) a novel sufficient condition
that can be checked automatically. We then take a step beyond
\invariantconfluence{} and introduce a generalization of \invariantconfluence{}
called segmented \invariantconfluence{}, which allows us to replicate
non-\invariantconfluent{} objects with a small amount of coordination.

### 2018-09-20 doc/paper/sections/interactive_decision_procedure.tex
Last tweaks!
Discussed limitations + shrunk paper.
From reviewer 11:

> Discussion of limitations and practical use can be made clearer.

From reviewer 12:

> While the ideas of the paper are very nice, the paper may more
> concretely describe how practical the algorithms are. For example, the
> first decision procedure (Algorithm 1) involves human to augment NR.
> In general, what are the tricky classes of transactions that may
> require lot of human interactions? Similar comments apply to segmented
> invariant confluence (Algorithm 5). What classes of
> transactions/applications that may req

\subsection{The Decision Procedure}
A user provides \algoref{InteractiveDecisionProcedure} with an object $O = (S,
\join)$, a start state $s_0$, a set of transactions $T$, and an invariant
$I$\footnote{%
  Formally, $I$ is a (possibly infinite) set of states (e.g. $\setst{(x, y) \in
  \ints \times \ints}{xy \leq 0}$), but in practice, our implementation of
  \IsInvConfluent{} takes in $I$ as a finite logical formula (e.g. $xy \leq
  0$).
}. The user then interacts with
\algoref{InteractiveDecisionProcedure} to iteratively eliminate unreachable
states from the invariant. Meanwhile, the algorithm leverages an
\invariantclosure{} decision procedure to either (a) conclude that $O$ is or is
not \sTIconfluent{} or (b) provide counterexamples to the user to help them
eliminate unreachable states. After all unreachable states have been eliminated
from the invariant, \thmref{IclosureEquivalentIconfluence} allows us to reduce
the problem of \invariantconfluence{} directly to the problem of
\invariantclosure{}, and the algorithm terminates.
%
We now describe \algoref{InteractiveDecisionProcedure} in detail. An example of
how to use \algoref{InteractiveDecisionProcedure} on \exampleref{Z2} is given
in \figref{DecisionProcedureExample}.

@@

\subsection{Tolerating User Error}
\algoref{InteractiveDecisionProcedure} is an \emph{interactive} decision
procedure. It requires that a user classify states as reachable or unreachable.
To err is human, so what happens when a user incorrectly classifies a state?
There are two possible errors that can be made, and
\algoref{InteractiveDecisionProcedure} can be made robust to both.

@@

\textbf{A user can label a reachable state as unreachable.}
In this case, \IsIclosed$(O, I - NR)$ might return true, even though $O$ is not
\sTIconfluent. Thus, a user might falsely believe their distributed object to
be \sTIconfluent{} even though it isn't, and eventually one replica of their
distributed object might enter a state that violates the invariant.
%
We can mitigate this in two ways. First, we can aggressively expand $R$ automatically. If a
user ever labels a state $s$ as unreachable, but $s \in R$, we can notify the
user of their mistake. Second, \Helper{} returns true when $O$ is \iclosed{(I -
NR)} for some $NR$, so $O$ \emph{is} \sticonfluent{s_0}{T}{I-NR} (even though
it might not be \sTIconfluent{}). Thus, when a user replicates their
distributed object across a set of servers, they can deploy with the invariant
$I - NR$ instead of $I$. If $NR$ is correct and only contains unreachable
states, then deploying with $I - NR$ is equivalent to deploying with $I$. If
$NR$ is incorrect and contains some \sTIreachable{} states, then some of these
states are artificially made unreachable, but the system is still guaranteed to
never produce a state which violates $I$.

>>>>

\subsection{The Decision Procedure}
A user provides \algoref{InteractiveDecisionProcedure} with an object $O = (S,
\join)$, a start state $s_0$, a set of transactions $T$, and an invariant
$I$\footnote{%
  Formally, $I$ is a (possibly infinite) set of states (e.g. $\setst{(x, y) \in
  \ints \times \ints}{xy \leq 0}$), but in practice, our implementation of
  \IsInvConfluent{} takes in $I$ as a finite logical formula (e.g., $xy \leq
  0$).
}. The user then interacts with
\algoref{InteractiveDecisionProcedure} to iteratively eliminate unreachable
states from the invariant. Meanwhile, the algorithm leverages an
\invariantclosure{} decision procedure to either (a) conclude that $O$ is or is
not \sTIconfluent{} or (b) provide counterexamples to the user to help them
eliminate unreachable states. After all unreachable states have been eliminated
from the invariant, \thmref{IclosureEquivalentIconfluence} allows us to reduce
the problem of \invariantconfluence{} directly to the problem of
\invariantclosure{}, and the algorithm terminates.
%
We now describe \algoref{InteractiveDecisionProcedure} in detail. An example of
how to use \algoref{InteractiveDecisionProcedure} on \exampleref{Z2} is given
in \figref{DecisionProcedureExample}.

@@

\subsection{Limitations}
Our interactive \invariantconfluence{} decision procedure has two main
limitations. First, \algoref{InteractiveDecisionProcedure} requires an
\invariantclosure{} decision procedure, but determining \invariantclosure{} is
undecidable in general.
%
\begin{techreport}

  For example, let $O_p = (S, \join)$ where $S$ is the set of all programs and
  $s_1 \join s_2 = p$ for some fixed program $p$. Letting $I$ be the set of all
  programs that terminate, determining if $O_p$ is \Iclosed{} is tantamount to
  determining if $p$ terminates.
\end{techreport}
%
In practice, we can implement an \invariantclosure{} decision procedure using
an SMT solver like Z3 that works well on simple objects, invariants, and merge
functions (e.g., integers, tuples, infinite sets, bit vectors, linear
constraints, basic arithmetic, tuple projection, basic set operations, bit
arithmetic). But, SMT solvers are mostly unable to analyze more complex
constructs (e.g., finite lists~\cite{kroning2009proposal}, graphs, recursive
algebraic data types, nonlinear constraints, merge functions that contain loops
or recursion).

Second, \algoref{InteractiveDecisionProcedure} relies on a user to identify
unreachable states. As we saw in \figref{DecisionProcedureExample}, the set of
unreachable states can sometimes be clear, especially if there's a noticeable
pattern in the set of reachable states. However, if the set of transactions is
large or complex or if the merge function is complex, then reasoning about
unreachable states can be difficult.  Unlike with reachable states---where
verifying that a state is reachable only requires thinking of a single way to
reach the state---verifying that a state is unreachable requires a programmer
to reason about a large number of system executions and conclude that
\emph{none} of them can lead to the state. In the future, we plan on exploring
ways to assist a user reason about unreachable states and discover some
unreachable states automatically.

\begin{techreport}
  \subsection{Tolerating User Error}
  \algoref{InteractiveDecisionProcedure} is an \emph{interactive} decision
  procedure. It requires that a user classify states as reachable or
  unreachable.  To err is human, so what happens when a user incorrectly
  classifies a state?  There are two possible errors that can be made, and
  \algoref{InteractiveDecisionProcedure} can be made robust to both.

@@

  \textbf{A user can label a reachable state as unreachable.}
  In this case, \IsIclosed$(O, I - NR)$ might return true, even though $O$ is
  not \sTIconfluent. Thus, a user might falsely believe their distributed
  object to be \sTIconfluent{} even though it isn't, and eventually one replica
  of their distributed object might enter a state that violates the invariant.
  %
  We can mitigate this in two ways. First, we can aggressively expand $R$
  automatically. If a user ever labels a state $s$ as unreachable, but $s \in
  R$, we can notify the user of their mistake. Second, \Helper{} returns true
  when $O$ is \iclosed{(I - NR)} for some $NR$, so $O$ \emph{is}
  \sticonfluent{s_0}{T}{I-NR} (even though it might not be \sTIconfluent{}).
  Thus, when a user replicates their distributed object across a set of
  servers, they can deploy with the invariant $I - NR$ instead of $I$. If $NR$
  is correct and only contains unreachable states, then deploying with $I - NR$
  is equivalent to deploying with $I$. If $NR$ is incorrect and contains some
  \sTIreachable{} states, then some of these states are artificially made
  unreachable, but the system is still guaranteed to never produce a state
  which violates $I$.
\end{techreport}

### 2018-09-20 doc/paper/sections/merge_reduction.tex
Discussed limitations + shrunk paper.
From reviewer 11:

> Discussion of limitations and practical use can be made clearer.

From reviewer 12:

> While the ideas of the paper are very nice, the paper may more
> concretely describe how practical the algorithms are. For example, the
> first decision procedure (Algorithm 1) involves human to augment NR.
> In general, what are the tricky classes of transactions that may
> require lot of human interactions? Similar comments apply to segmented
> invariant confluence (Algorithm 5). What classes of
> transactions/applications that may require complex

The proof of \thmref{ReducibilityImpliesIconfluence} is a straightforward
induction. We begin with an \sTIreachable{} expression $e$ and repeatedly
replace any subexpression that merges two merge-free subexpressions with an
equivalent merge-free reachable subexpression (which we can do because $O$ is
merge-reducible). We repeat this process until $e$ has been completely replaced
with an equivalent merge-free reachable expression $e'$. Because $I(s_0)$ and
because our system model only executes transactions that preserve the
invariant, $e'$ (and hence $e$) is guaranteed to satisfy the invariant. Thus,
all reachable states satisfy the invariant, so $O$ is \invariantconfluent{}.

@@

The proof of \thmref{LatticeProperty} involves proving that if criteria (1)
through (4) are met, then $O$ is \sTImergereducible{}. From this, it's
immediate that $O$ is \sTIconfluent{} by
\thmref{ReducibilityImpliesIconfluence}. To show that $O$ is
\sTImergereducible, we consider two merge-free reachable expressions
\[
  e_1 = t_1(\ldots(t_n(s_0))\ldots)
  \quad\text{and}\quad
  e_2 = u_1(\ldots(u_m(s_0))\ldots)
\]
Then, we inductively use criterion (3) with criteria (1) and (2) to show that
\[
  e_3 = t_1(\ldots(
          t_n(u_1(\ldots(u_m(s_0))\ldots))
        )\ldots)
\]
is reachable and evaluates to $e_1 \join e_2$.

@@

  Clearly, criteria (1), (2), and (4) are satisfied. Criterion (3) is also
  satisfied. If $s$ is an arbitrary state that satisfies $I$, then it is
  non-empty and contains, without loss of generality, only even integers. If
  $t_1$ and $t_2$ are arbitrary transactions such that $I(t_1(s))$ and
  $I(t_2(s))$, then $t_1(s)$ and $t_1(s)$ are also non-empty and contain only
  even integers. Thus, $t_1(s) \cup t_2(s)$ is clearly non-empty and contains
  only even integers, so $I(t_1(s) \join t_2(s))$.
  %
  However, $O$ is not \Iclosed{}. Let $s_1 = \set{0}$ and $s_2 = \set{1}$.
  Then, $I(s_1)$ and $I(s_2)$, but letting $s_3 = s_1 \cup s_2 = \set{0, 1}$,
  $\lnot I(s_3)$.
\end{example}

>>>>

\begin{techreport}
  The proof of \thmref{ReducibilityImpliesIconfluence} is a straightforward
  induction. We begin with an \sTIreachable{} expression $e$ and repeatedly
  replace any subexpression that merges two merge-free subexpressions with an
  equivalent merge-free reachable subexpression (which we can do because $O$ is
  merge-reducible). We repeat this process until $e$ has been completely
  replaced with an equivalent merge-free reachable expression $e'$. Because
  $I(s_0)$ and because our system model only executes transactions that
  preserve the invariant, $e'$ (and hence $e$) is guaranteed to satisfy the
  invariant. Thus, all reachable states satisfy the invariant, so $O$ is
  \invariantconfluent{}.
\end{techreport}

@@

\begin{techreport}
  The proof of \thmref{LatticeProperty} involves proving that if criteria (1)
  through (4) are met, then $O$ is \sTImergereducible{}. From this, it's
  immediate that $O$ is \sTIconfluent{} by
  \thmref{ReducibilityImpliesIconfluence}. To show that $O$ is
  \sTImergereducible, we consider two merge-free reachable expressions
  \[
    e_1 = t_1(\ldots(t_n(s_0))\ldots)
    \quad\text{and}\quad
    e_2 = u_1(\ldots(u_m(s_0))\ldots)
  \]
  Then, we inductively use criterion (3) with criteria (1) and (2) to show that
  \[
    e_3 = t_1(\ldots(
    t_n(u_1(\ldots(u_m(s_0))\ldots))
    )\ldots)
  \]
  is reachable and evaluates to $e_1 \join e_2$.
\end{techreport}

@@

  Criteria (1), (2), (3) and (4) are all satisfied. However, $O$ is not
  \Iclosed{}. Let $s_1 = \set{0}$ and $s_2 = \set{1}$. Then, $I(s_1)$ and
  $I(s_2)$, but letting $s_3 = s_1 \cup s_2 = \set{0, 1}$, $\lnot I(s_3)$.

  \begin{techreport}

    Here's why criterion (3) is satisfied. If $s$ is an arbitrary state that
    satisfies $I$, then it is non-empty and contains, without loss of generality,
    only even integers. If $t_1$ and $t_2$ are arbitrary transactions such that
    $I(t_1(s))$ and $I(t_2(s))$, then $t_1(s)$ and $t_1(s)$ are also non-empty
    and contain only even integers. Thus, $t_1(s) \cup t_2(s)$ is clearly
    non-empty and contains only even integers, so $I(t_1(s) \join t_2(s))$.
  \end{techreport}
\end{example}
